Stateful Forward Reasoning

### 2025-02-04 paper.tex
Add missing csquotes package import
Typo
Fix semicolons in tikzpictures
Some implementation content

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage{newunicodechar}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{hyperref}

@@

\newunicodechar{₁}{\ensuremath{_{1}}}
\newunicodechar{₂}{\ensuremath{_{2}}}
\newunicodechar{₃}{\ensuremath{_{3}}}
\newunicodechar{₄}{\ensuremath{_{4}}}
\newunicodechar{₅}{\ensuremath{_{5}}}
\newunicodechar{₆}{\ensuremath{_{6}}}
\newunicodechar{₇}{\ensuremath{_{7}}}
\newunicodechar{₈}{\ensuremath{_{8}}}
\newunicodechar{₉}{\ensuremath{_{9}}}
\newunicodechar{ₙ}{\ensuremath{_{n}}}
\newunicodechar{ₘ}{\ensuremath{_{m}}}
\newunicodechar{ᵢ}{\ensuremath{_{i}}}
\newunicodechar{ⱼ}{\ensuremath{_{j}}}
\newunicodechar{ₗ}{\ensuremath{_{l}}}
\newunicodechar{ₖ}{\ensuremath{_{k}}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Φ}{\ensuremath{\Phi}}
\newunicodechar{Ψ}{\ensuremath{\Psi}}
\newunicodechar{Ω}{\ensuremath{\Omega}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{∉}{\ensuremath{\notin}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\nequiv}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≔}{\ensuremath{\coloneq}}
\newunicodechar{∧}{\ensuremath{\land}}
\newunicodechar{∨}{\ensuremath{\lor}}
\newunicodechar{∪}{\ensuremath{\cup}}
\newunicodechar{∩}{\ensuremath{\cap}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{↦}{\ensuremath{\mapsto}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{∷}{\ensuremath{\mathbin{::}}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{⊤}{\ensuremath{\top}}
\newunicodechar{⊥}{\ensuremath{\bot}}

@@

@@

\begin{center}
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
  \node[xshift = 0] {$\eta_y$}
    child[xshift = -20] {node {1}
      child {node {$302$}
        child {node {$\{h_1\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
        }
    child[xshift = 20] {node {2}
      child {node {$301$}
        child {node {$\{h_1\}$}}}
    %  child {node {$300$}
    %    child {node {$\{\}$}}}
    };
  \node[xshift=160] {$\mu_y$}
    child {node[xshift=-20] {1}
      child {node {$302$}
        child {node {$\{[h_1]\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
    }
    %child {node[xshift=20] {2}
    %  child {node {$301$}
    %    child {node[yshift=-2em] {$\{\}$}}}};
\end{tikzpicture}
\end{center}

@@

% NOTE : remove comments to get full trees.
\begin{tikzpicture}[level distance=12mm, scale=0.50]
    \node{}
    \node[xshift = 30] {$\eta_x$}
        %child[xshift = -20] {node {1}
        %  child {node {$a$}
        %    child {node {$\{ \}$}}}
        %  child {node {$b$}
        %    child {node {$\{ \}$}}}}
        child[xshift = 40] {node {2}
        child {node {$a$}
            child {node {$\{h_1\}$}}}
        % child {node {$b$}
        %    child {node {$\{ \}$}}}
        };
    \node[xshift = 110] {$\eta_y$}
        %child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}}
        child[xshift = 40] {node {2}
            child {node {$a$}
                child {node {$\{h_1\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        };
    \node[xshift = 190, yshift = -25] {$\mu_x$}
        %child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}}
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
    \node[xshift = 270, yshift = -25] {$\mu_y$}
        %child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}}
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
\end{tikzpicture}

@@

% NOTE : uncomment to get full trees.
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
    \node[xshift = 0] {$\eta_x$}
        child[xshift = -20] {node {1}
            child {node {$a$}
                child {node {$\{h_2\}$}}}
            %child {node {$b$}
            %    child {node {$\{ \}$}}}
            }
        child[xshift = 20] {node {2}
            child {node {$a$}
                child {node {$\{h_1\}$}}}
            %child {node {$b$}
            %    child {node {$\{ \}$}}}
            };
    \node[xshift = 80] {$\eta_y$}
        child[xshift = -20] {node {1}
            %child {node {$a$}
            %    child {node {$\{ \}$}}}
            child {node {$b$}
                child {node {$\{ h_2\}$}}}}
        child[xshift = 20] {node {2}
            child {node {$a$}
                child {node {$\{h_1\}$}}}
            %child {node {$b$}
            %    child {node {$\{ \}$}}}
                };
    \node[xshift = 160, yshift = 0] {$\mu_x$}
        child[xshift = -20] {node {1}
            child {node {$a$}
                child {node {$\{[h_2]\}$}}}
            %child {node {$b$}
            %   child {node {$\{ \}$}}}
                }
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
    \node[xshift = 240, yshift = 0] {$\mu_y$}
        child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
            child {node {$b$}
                child {node {$\{[h_2]\}$}}}
        }
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
\end{tikzpicture}

@@

We are now ready to look for potential matches that could be extended by $h_3$.
In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{x,y\}$.
Then,
\[
\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right) = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\}
\]

@@

@@

\xcom{Here are some notes about the implementation.}
The index used to match hypotheses with their matching rules are implemented as discrimination trees.
A discrimination tree is a trie-like data structure that maps expressions to arbitrary data.
In~\cite{TODO}, it is shown to be good solution for the problem of finding terms (here the types of hypotheses) that might unify with other terms (here the types of the input premises).
In Lean 4, discrimination trees are also used to index typeclass instances and simplifier lemmas.

@@

The maps described in the last section are all implemented as persistent hash maps.
\xcom{Why did we not use discrimination trees? $+$ (pers. important cause we store related forward state for many goal.)}

One of the more costly operation that we need to run multiple times is the unification \xcom{defeq?} between the hypotheses and the types of the input premises associated to rules.
We implement a variant of Lean's unification process as follows: the idea is that we wish to normalize the types as much as possible.

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage{newunicodechar}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{hyperref}
\usepackage{csquotes}

@@

\newunicodechar{₁}{\ensuremath{_{1}}}
\newunicodechar{₂}{\ensuremath{_{2}}}
\newunicodechar{₃}{\ensuremath{_{3}}}
\newunicodechar{₄}{\ensuremath{_{4}}}
\newunicodechar{₅}{\ensuremath{_{5}}}
\newunicodechar{₆}{\ensuremath{_{6}}}
\newunicodechar{₇}{\ensuremath{_{7}}}
\newunicodechar{₈}{\ensuremath{_{8}}}
\newunicodechar{₉}{\ensuremath{_{9}}}
\newunicodechar{ₙ}{\ensuremath{_{n}}}
\newunicodechar{ₘ}{\ensuremath{_{m}}}
\newunicodechar{ᵢ}{\ensuremath{_{i}}}
\newunicodechar{ⱼ}{\ensuremath{_{j}}}
\newunicodechar{ₗ}{\ensuremath{_{l}}}
\newunicodechar{ₖ}{\ensuremath{_{k}}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{η}{\ensuremath{\eta}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Φ}{\ensuremath{\Phi}}
\newunicodechar{Ψ}{\ensuremath{\Psi}}
\newunicodechar{Ω}{\ensuremath{\Omega}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{∉}{\ensuremath{\notin}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\nequiv}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≔}{\ensuremath{\coloneq}}
\newunicodechar{∧}{\ensuremath{\land}}
\newunicodechar{∨}{\ensuremath{\lor}}
\newunicodechar{∪}{\ensuremath{\cup}}
\newunicodechar{∩}{\ensuremath{\cap}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{↦}{\ensuremath{\mapsto}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{∷}{\ensuremath{\mathbin{::}}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{⊤}{\ensuremath{\top}}
\newunicodechar{⊥}{\ensuremath{\bot}}

@@

\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
\newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
\newcommand{\mvar}[1]{\ensuremath{?#1}}
\newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}

@@

\begin{center}
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
  \node[xshift = 0] {$\eta_y$}
    child[xshift = -20] {node {1}
      child {node {$302$}
        child {node {$\{h_1\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
        }
    child[xshift = 20] {node {2}
      child {node {$301$}
        child {node {$\{h_1\}$}}}
    %  child {node {$300$}
    %    child {node {$\{\}$}}}
    };
  \node[xshift=160] {$\mu_y$}
    child {node[xshift=-20] {1}
      child {node {$302$}
        child {node {$\{[h_1]\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
    };
    %child {node[xshift=20] {2}
    %  child {node {$301$}
    %    child {node[yshift=-2em] {$\{\}$}}}};
\end{tikzpicture}
\end{center}

@@

% NOTE : remove comments to get full trees.
\begin{tikzpicture}[level distance=12mm, scale=0.50]
    \node[xshift = 30] {$\eta_x$}
        %child[xshift = -20] {node {1}
        %  child {node {$a$}
        %    child {node {$\{ \}$}}}
        %  child {node {$b$}
        %    child {node {$\{ \}$}}}}
        child[xshift = 40] {node {2}
        child {node {$a$}
            child {node {$\{h_1\}$}}}
        % child {node {$b$}
        %    child {node {$\{ \}$}}}
        };
    \node[xshift = 110] {$\eta_y$}
        %child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}}
        child[xshift = 40] {node {2}
            child {node {$a$}
                child {node {$\{h_1\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        };
    \node[xshift = 190, yshift = -25] {$\mu_x$};
        %child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}}
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
    \node[xshift = 270, yshift = -25] {$\mu_y$};
        %child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}}
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
\end{tikzpicture}

@@

% NOTE : uncomment to get full trees.
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
    \node[xshift = 0] {$\eta_x$}
        child[xshift = -20] {node {1}
            child {node {$a$}
                child {node {$\{h_2\}$}}}
            %child {node {$b$}
            %    child {node {$\{ \}$}}}
            }
        child[xshift = 20] {node {2}
            child {node {$a$}
                child {node {$\{h_1\}$}}}
            %child {node {$b$}
            %    child {node {$\{ \}$}}}
            };
    \node[xshift = 80] {$\eta_y$}
        child[xshift = -20] {node {1}
            %child {node {$a$}
            %    child {node {$\{ \}$}}}
            child {node {$b$}
                child {node {$\{ h_2\}$}}}}
        child[xshift = 20] {node {2}
            child {node {$a$}
                child {node {$\{h_1\}$}}}
            %child {node {$b$}
            %    child {node {$\{ \}$}}}
                };
    \node[xshift = 160, yshift = 0] {$\mu_x$}
        child[xshift = -20] {node {1}
            child {node {$a$}
                child {node {$\{[h_2]\}$}}}
            %child {node {$b$}
            %   child {node {$\{ \}$}}}
                };
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
    \node[xshift = 240, yshift = 0] {$\mu_y$}
        child[xshift = -20] {node {1}
        %    child {node {$a$}
        %        child {node {$\{ \}$}}}
            child {node {$b$}
                child {node {$\{[h_2]\}$}}}
        };
        %child[xshift = 20] {node {2}
        %    child {node {$a$}
        %        child {node {$\{\}$}}}
        %    child {node {$b$}
        %        child {node {$\{ \}$}}}
        %};
\end{tikzpicture}

@@

We are now ready to look for potential matches that could be extended by $h_3$.
In this particular case, since $h_3$ is in slot $2$ we look for $s(2) = \{x,y\}$.
Then,
\[
\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right) = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\}
\]

@@

\jcom{TODO: deletion}

@@

Implementing the preceding technique efficiently requires some care, particularly in the context of dependent type theory.
We briefly discuss the most prominent challenges.

\subsection{Unification}

In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
As a result, unifying premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
Our technique therefore seeks to minimise the number of unifications performed.

To limit the expense of unification, we perform unification at \emph{reducible} transparency.
Lean has multiple transparency levels that determine which constants are unfolded.
\enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions, which greatly reduces the cost of unification.
It also limits the power of forward reasoning since

We also need to account for higher-order unification problems: $Q~\mvar{f}$, where $Q$ is a predicate over functions and $\mvar{f}$ is a function variable, should unify with $Q~(\Lam{x}{x})$.
Since higher-order unification is undecidable and does not admit most general unifiers (TODO ref), we use instead a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are at least well-acquainted with this source of incompleteness.

\subsection{Data Representation}

Since the maps $μ_{x}$  and $η_{x}$ have the same domain, we may combine them into one map.
This core data structure is then represented by nested persistent hash maps: a first hash map maps variables to a second hash map, the second hash map maps slots to a third hash map and the third hash map maps instantiations to a set of hypotheses and a set of partial matches.

When we compare instantiations, the comparison should be up to definitional equality.
For example, when we try to apply a theorem $\All{x}{P~x → Q~x → R~x}$ where the first slot matches a hypothesis of type $P~a$ and the second slot matches a hypothesis of type $P~((\Lam{x}{x})~a)$, the application should succeed even though the instantiations of $\mvar{x}$ are different.
We therefore do not store $a$ and $(\Lam{x}{x})~a$ in the variable maps as-is.

Instead, we bring all instantiations into a normal form which we call \emph{reducible proof-irrelevant normal form} (RPINF).
This is the usual normal form of expressions with respect to Lean's notion of computation, except for two modifications:
\begin{itemize}
  \item Only \emph{reducible} defined constants are unfolded, i.e.\ replaced with their definiendum.
        This reflects our choice to perform unification at reducible transparency.
  \item Parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
        This reflects the fact that Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
        When computing the RPINF, we also mark all proofs occurring in it with some special \enquote{metadata}.
\end{itemize}

With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if their RPINFs, $t'$ and $u'$, are syntactically equal except for those portions of $t'$ and $u'$ that are proofs, which are always considered equal.
Since this is an entirely syntactic notion of equality, it can be checked very quickly and we can define an effective hash function that is compatible with it.
This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.

We believe that this scheme is more efficient than alternative schemes relying on repeated unification.
For example, we could use Lean's implementation of imperfect discrimination trees (TODO ref), which implement a map from expressions to arbitrary data where lookup respects definitional equality.
However, since these discrimination trees are imperfect, they may return spurious lookup results, so whether a lookup result is valid must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
Repeatedly performing such unifications may be more expensive than calculation one RPINF per expression (which can also be easily cached).

\subsection{Indexing}

\subsection{Aesop Integration}

\xcom{Here are some notes about the implementation.}
The index used to match hypotheses with their matching rules are implemented as discrimination trees.

@@
